Generalized Linear Mixed Models

### 2009-04-03 docs/inla/inla.rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@177 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\newcommand{\comment}[1]{}

@@

\begin{frame}[fragile]
\frametitle{Parameter estimates}
<<bacpar,fig=false,echo=true>>=
bacSummary = summaryChain(bacParams)
bacSummary$scalars[,c(1:3,5,8)]
bacSummary$RID[1:4,c(1:3,5,8)]

@
\end{frame}

@@

\comment(

\comment{
	jpeg('Figures/ontarioCancer.jpg', height=600, width=800)
}

@@

\begin{frame}
\frametitle{Integrated Nested Laplace Approximations}

\begin{frame}
\frametitle{How it works}

\begin{frame}
\frametitle{Example: }
<<bacteriaINlA,fig=false>>=
#library(MASS)
#data(bacteria)
#bacterianew <- bacteria[,-(2:3)]
#bacterianew$y = as.integer(bacteria$y=="y")
#levels(bacterianew$trt) <- c("placebo", "drug", "drugplus")
#bacteria=bacterianew
# do this pql = glmmPQL(y ~ trt, random = ~1|ID, family='binomial',data=bacteria)

@

\begin{frame}
\frametitle{Example: Lung cancer in Ontario}
<<cancerInla,fig=false>>=
if(F) {
library(diseasemapping)
library(sp)
data(popdata)
data(casedata)
model = getRates(casedata, popdata, ~age*sex)
ontario = getSMR(popdata,model,  casedata)
data(popDataAdjMat)

# fit bym model with inla, like
#forBugs = glmmBUGS(formula=Observed + logExpected ~ 1,
#  effects="CSDUID",   family="poisson", spatial=popDataAdjMat,
#  data=ontario)
}
@
\end{frame}

>>>>

\def\mm#1{\ensuremath{\boldsymbol{#1}}} % version: amsmath
\newcommand{\comment}[1]{}

@@

\begin{frame}[fragile]
\frametitle{Parameter estimates}
<<bacpar,fig=false,echo=true>>=
bacSummary = summaryChain(bacParams)
bacSummary$scalars[,c(1:3,5,8)]
bacSummary$betas[,c(1:3,5,8)]
bacSummary$RID[1:4,c(1:3,5,8)]
#$
@
\end{frame}

@@

\comment{
	jpeg('Figures/ontarioCancer.jpg', height=600, width=800)

}

@@

\begin{frame}
\frametitle{Integrated Nested Laplace Approximations}
\begin{itemize}
\item Notice that
\[
\[
\pi[\theta|Y] = \frac{\pi[U,\theta|Y]}{\pi[U|\theta,Y]}
\]
\item Laplace approximation
\[
\tilde\pi[\theta|Y] \propto \left. \frac{\pi[U,\theta|Y]}{\tilde\pi_G[U|\theta, Y]} \right| \theta=\theta^*
\]
where $\theta^*$ is the mode, and $\pi_G$ is a Gaussian approximation.
	\item \[
	\pi[U,\theta|Y] \propto \pi[\theta]\pi[U|\theta] \prod_{i} \pi[Y_i | U_i,\theta]
	\]

\end{itemize}
\end{frame}

\begin{frame}
\begin{itemize}
	\item Do the following integrals numerically
\end{itemize}
\frametitle{Marginal posteriors}
           \begin{align}
                \widetilde{\pi}(U_{i}\mid\mm{y}) &= \int
                \widetilde{\pi}(\mm{\theta} \mid \mm{y}) \;
                \widetilde{\pi}(U_{i} \mid \mm{\theta}, \mm{y})\; d\mm{\theta}\nonumber\\
                \widetilde{\pi}(\theta_{j} \mid \mm{y}) &= \int
                \widetilde{\pi}(\mm{\theta} \mid \mm{y}) \;
                d\mm{\theta}_{-j} \nonumber
            \end{align}
\end{frame}

\begin{frame}
    \frametitle{Remarks}
    \begin{enumerate}
    \item Expect $\widetilde{\pi}(\mm{\theta} | \mm{y})$ to be
        accurate, since
        \begin{itemize}
        \item $\mm{x}|\mm{\theta}$ is \emph{a priori} Gaussian
        \item Likelihood models are `well-behaved' so
            \begin{displaymath}
                \pi(\mm{x} \mid \mm{\theta}, \mm{y})
            \end{displaymath}
            is \emph{almost} Gaussian.
        \end{itemize}
    \item There are no distributional assumptions on $\mm{\theta} |
        \mm{y}$
    \item Similar remarks are valid to
        \begin{displaymath}
            \widetilde{\pi}(x_{i}\mid\mm{\theta}, \mm{y})
        \end{displaymath}
    \end{enumerate}
\end{frame}

\begin{frame}
\frametitle{How it works}
\begin{enumerate}
	\item Explore the space to find the model of $[\theta|Y]$
	\item Then compute the Laplace approximation for $[U|Y,\theta]$
	\item Then numerically integrate out the $\theta_i$.
	\item Use polynomials in place of the Gaussian to approximate the $[U_i|\theta,Y]$.
\end{enumerate}
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: bacteria}
{\small
\begin{verbatim}
bacteria$r<-bacteria$y=="y"
bacteria$n<-1
formula1<- r ~ trt + f(ID,model="iid",param=c(0.001,0.001))
mod1 = inla(formula1,family="binomial",data=bacteria,Ntrials=n)
summary(mod1) #This will give summaries of the posteriors

Fixed effects:
                 mean        sd 0.025quant 0.975quant    kld dist
(Intercept)  2.2955447 0.4747077   1.498364  3.3689509 0.155881406
trtdrug     -1.2147419 0.5831969  -2.442864 -0.1223520 0.027425825
trtdrug+    -0.7292562 0.5917214  -1.953683  0.3977671 0.008737907

Model hyperparameters:
                mean   sd     0.025quant 0.975quant
Precision for ID 0.9724 0.5651 0.2943     2.4230  
Expected number of effective parameters(std dev) :18.03(7.335)
Number of equivalent replicates : 12.20
\end{verbatim}
}
\end{frame}

\begin{frame}
\frametitle{Posterior distribution of precision}
\includegraphics[width=0.7\textwidth]{Prec_BACTERIA.pdf}
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: Lung cancer in Ontario}
{\small
\begin{verbatim}
nb2inla("ontario.graph",popDataAdjMat)
data$observed[is.na(data$logExpected)]<-NA
data$logExpected[is.na(data$logExpected)]<- -100 
data$region<-0:(length(data$CSDUID)-1)
data$region.spatial<-data$region

formula2<-observed ~ f(region.spatial,
    model="besag", graph.file="ontario.graph", 
    param=c(20,15),initial=2) + 
  f(region,model="iid", param=c(1,0.01))

mod2 <- inla(formula2,family="poisson",data=data,
  offset=logExpected,verbose=TRUE)

data$V<-mod2$summary.random$region[,2]
data$U<-mod2$summary.random$region.spatial[,2]

writePolyShape(ontario, "ontario")

\end{verbatim}
}
\end{frame}

\begin{frame}
\frametitle{Smoothed cancer risk}
\includegraphics{Map_LINEAR.pdf}
\end{frame}

\begin{frame}[fragile]
\frametitle{Results}
{\small

\begin{verbatim}
summary(mod2)

Fixed effects:
                mean       sd 0.025quant 0.975quant     kld dist
(Intercept) 0.7209629 3.654778  -6.456466   7.888327 7.134535e-05

Random effects:
Name      Model         Max KLD
region.spatial   Besags ICAR model   0.00083
region   IID model   0.01849

Model hyperparameters:
                            mean   sd     0.025quant 0.975quant
Precision for region.spatial 2.2255 0.3251 1.6469     2.9213  
Precision for region         1.9612 0.1960 1.6005     2.3700  
Expected number of effective parameters(std dev) :357.87(2.378)
Number of equivalent replicates : 1.115
\end{verbatim}

}
\end{frame}

\begin{frame}
\frametitle{Advantages/Disadvantanges}
\begin{block}{Advantages}
\begin{itemize}
	\item It appears that this is a very good approximation, and is much less computationally intensive than MCMC
	\item the code works in parallel, so is very fast on multi-core machines
\end{itemize}
\end{block}
\begin{block}{Disadvantages}
\begin{itemize}
\item Only marginals, not joint posteriors, are produced.
\item INLA is harder to modify than MCMC routines, though that might be only due to familiarity with MCMC
\item There are some things which INLA can't do, though in theory MCMC can do anything.
\item Starting values do have an effect, so INLA isn't fool-proof.
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Conclusions:}
\begin{itemize}
	\item There's no excuse for avoiding GLMM's
	\item or using PQL 
	\item Good models and good tools are available for dependent non-Gaussian data.
\end{itemize}
\end{frame}
